Finally tagless, partially evaluated

### 2008-02-27 tagless/talk.tex
Final talk at APLAS 2007

\pagenumbering{arabic}
\documentclass[ucs,professionalfont]{beamer}
\usepackage{concmath}
\usepackage{helvet}
%\renewcommand{\rmdefault}{cmss}
\usepackage[utf8x]{inputenc}
\usepackage[override]{cmtt}
\usepackage{mdwtab}
\usepackage{prooftree1} \proofrulebaseline=1.5ex \proofdotseparation=.75ex
\usepackage{microtype}
\usepackage{calc}

@@

% To get the page numbers to stay constant on a frame [dpt]
\mode<presentation>
{
  \setbeamertemplate{sidebar left}{\thispdfpagelabel{\insertframenumber}}
}

@@

@@

\onslide<1->
\begin{tabular}{@{}Mll@{}}
\only<5-8,11-14>{\text{\textcolor{structure.fg}{Object term}}}%
\only<16>{\rlap{\textcolor{structure.fg}{ML higher-order functor}}}
& \only<5-8,11-15>{\textcolor{structure.fg}{\only<-14>{\llap{\tikz\draw[->](0,0)--(1.5,0)node[anchor=mid,inner sep=0pt]{\vphantom{O}};\:}}\only<3-8>{Haskell term}\only<9-15>{ML higher-order functor}}} \\
\multicolumn{2}{@{}l@{}}{\only<16>{\alert{\texttt{module POWER (S:Symantics) = struct open S}}}} \\
\only<16>{\rlap{\alert{\texttt{ let term () =}}}}%
\visible<1-14>{\uncover<1-2,5-8,11-14>{\alert<6,12>{\fun{x} \fix{f} \fun{n}}}}
& \ttfamily
  \only<5-8>{\alert<6>{lam (\textbackslash x -> fix (\textbackslash f -> lam (\textbackslash n ->}}%
  \only<11-16>{\alert<12>{lam\,(fun x\,-> fix\,(fun f\,-> lam\,(fun n\,->}} \\
\visible<1-14>{\uncover<1-2,5-8,11-14>{\alert<8,14>{\cond{\textcolor{black}{n\le\alert<5,11>{0}}}{\textcolor{black}{1}}}}}
& \ttfamily
  \only<5-8>{\alert<8>{if\textunderscore} \alert<8>(leq n (\alert<5>{int 0})\alert<8>) \alert<8>(int 1\alert<8>)}%
  \only<11-16>{\alert<14>{if\textunderscore} \alert<14>(leq n (\alert<11>{int 0})\alert<14>) \alert<14>{(fun\,()\,->}\,int 1\alert<14>)} \\
\visible<1-14>{\uncover<1-2,5-8,11-14>{x\times \alert<7,13>{f(n-1)}}}
& \ttfamily
  \only<5-8>{\alert<8>(mul x (\alert<7>{app f (add n (int (-1)))})\alert<8>))))}%
  \only<11-16>{\alert<14>{(fun\,()\,->}\,mul x\,(\alert<13>{app f\,(add n\,(int\,(-1)\!)\!)}\!)\!\alert<14>)\!)\!)\!)} \\
\multicolumn{2}{@{}l@{}}{\only<16>{\alert{\texttt{end:\ functor (S:Symantics) -> sig}}}} \\
\only<16>{\rlap{\alert{\texttt{ val term:\ unit\,->}}}}%
\visible<1-14>{\uncover<1-2,5-8,11-14>{\mathord:\;\ZZ\to\ZZ\to\ZZ}}
& \ttfamily
  \only<5-8>{:\!:\ \!Symantics \!repr \!=>\,repr\,(Int\,->\,Int\,->\,Int)}%
  \only<11-16>{('c,\,int\,->\,int\,->\,int)\,\only<16>{\alert{S.}}repr} \\
\multicolumn{2}{@{}l@{}}{\only<16>{\alert{\texttt{end}}}}
\end{tabular}
\end{frame}

@@

\begin{frame}[fragile]{\only<11->{Type-indexed types}\only<13->{: }\only<-10,13-15>{Partial evaluation}\only<16->{CPS transformation}}
\setbeamercovered{transparent=20}
\newcommand\remember[1]{\tikz[remember picture,baseline=0pt,text=structure.fg] \node[inner ysep=0pt,inner xsep=2pt,anchor=base] (#1) {#1};}
\newcommand\remark[1]{\textcolor{structure.fg}{\footnotesize(* #1 *)}}
\definecolor{dynamic}{rgb}{.7,0,.7}
\definecolor{static}{rgb}{0,.7,.7}
\newcommand\dynamic[1]{\textcolor{dynamic}{\texttt{‹#1›}}}
\newcommand\static[1]{\textcolor{static}{\texttt{#1}}}
\begin{overlayarea}{0pt}{3ex}
\onslide+<1,13-15>{\rlap{\hspace*{-1pc}\texttt{module P = struct}}}%
\onslide+<12>\vspace*{-1em}%
\begin{semiverbatim}
module type Symantics = sig type ('c,\alert<12>{'s},'a) repr
 val int: int -> ('c,\alert<12>{int},int) repr
 val lam: 'x\,-> ('c, \alert<12>{('c,'s,'a) repr ->
            \,        ('c,'t,'b) repr} as\,'x, 'a\,->\,'b) repr
\uncover<0>{ val fix: (('c, \alert<12>{('c,'s,'a) repr -> ('c,'t,'b) repr},
                'a -> 'b) repr as 'x -> 'x) -> 'x}
 val app: ('c, \alert<12>{('c,'s,'a) repr ->
               ('c,'t,'b) repr} as\,'x, 'a\,->\,'b) repr ->\,'x
\uncover<0>{ val add: 'x -> 'x -> (('c,\alert<12>{int},int) repr as\,'x)
 val if_: ('c,\alert<12>{bool},bool) repr -> (unit->'x) -> (unit->'x)
                              -> (('c,\alert<12>{'s},'a) repr as\,'x)}
end
\end{semiverbatim}
\end{overlayarea}
\begin{tabular}[L]{@{}>{\ttfamily\frenchspacing}l@{\hspace*{-.5em}}r@{ }cc@{}}
\onslide+<1-10,13->
type ('c,\alt<13->{\only<14->{\alert<14-15>{'s},}'a}{\alt<1>{ 'a}{\alert<2>{int}}}) repr
                    & \onslide+<3-10>\remember{source term}
                    & \onslide+<4-10>$3$
                    & \onslide+<5-10>$x$ \\
\onslide+<1-10,13->
\ \ \ = \alt<13->{\alt<16->{\makebox[14pc][l]{\alert{('s\,->\,ans)\,->\,ans}}\remark{CBN CPS evaluator}}{('c,'a) code}}
                 {\alt<1>{???}{('c,int) code}}
                    & \onslide+<3-10>\remember{dynamic part}
                    & \onslide+<4-10>\dynamic{3}
                    & \onslide+<5-10>\dynamic{x} \\
\onslide+<2-10,13->
\ \ \ \alt<16->{= \makebox[14pc][l]{\alert{('c, ('s\,->\,ans)\,->\,ans) code}}\remark{CBN CPS compiler}}{* \alt<13->{\alert{\alt<14->{'s}{('c,'a) static}}}{int} option}
                    & \onslide+<3-10>\remember{static part}
                    & \onslide+<4-10>\static{Some 3}
                    & \onslide+<5-10>\static{None} \\
\onslide+<14-15>...\\
\onslide+<14-15>\rlap{\hspace*{-1pc}end}%
\onslide+<6-10>
type ('c,\alert<6>{int->int}) repr
                    && \onslide+<8-10>\hspace*{-4em}$\fun{x}x$
                    & \onslide+<7-10>$f$ \\
\onslide+<6-10>
\ \ \ = ('c,int->int) code
                    && \onslide+<8-10>\hspace*{-4em}\dynamic{fun \!x->x}
                    & \onslide+<7-10>\dynamic{f} \\
\onslide+<15>\rlap{\hspace*{-1pc}module POWER7P = POWER7(P)}%
\onslide+<6-10>
\ \ \ * \alert<8>{(('c,int) repr ->}
                    && \onslide+<8-10>\hspace*{-4em}\static{Some (fun \!r->r)}
                    & \onslide+<7-10>\static{None} \\
\onslide+<15>\rlap{\hspace*{-1pc}\>POWER7P.term ()}%
\onslide+<6-10>
\ \ \ \ \ \ \alert<8>{('c,int) repr)} option \\
\onslide+<15>\rlap{\hspace*{-1pc}\ \ (‹fun x -> x\,*\,x\,*\,x\,*\,x\,*\,x\,*\,x\,*\,x›, Some <fun>)}%
\\
\onslide+<9-10>
type ('c,'a) \\
\ \ \ = ('c,'a) code \\
\ \ \ * \alert<9-10>{\alt<9>{???}{('c,'a) static}} option
\end{tabular}
\onslide+<10-13,14->
\texttt{\begin{tabular}[L]{@{}l@{ }l@{}}
    type ('c,\:int)        &static = int \\
    type ('c,\:bool)       &static = bool \\
    type ('c,\:'a\,->\,'b) &static = ('c,'a)\:repr -> ('c,'b)\:repr
\end{tabular}}

@@

\begin{frame}{Conclusion}
    Early obvious safety

    Initial type-checking can be done

    Accommodates interpretations with type-indexed types

    Preserves sharing

    Acknowledge prior-work authors in slides
\end{frame}

>>>>

\pagenumbering{arabic}
\documentclass[ucs,professionalfont]{beamer}
\usepackage{concmath}
\usepackage{helvet}
%\renewcommand{\rmdefault}{cmss}
\usepackage[utf8x]{inputenc}
\usepackage[override]{cmtt}
\usepackage{mdwtab}
\usepackage{prooftree1} \proofrulebaseline=1.5ex \proofdotseparation=.75ex
\usepackage{calc}

@@

% Record frame numbers as PDF page labels
\mode<presentation>
{
  \setbeamertemplate{sidebar left}{\thispdfpagelabel{\insertframenumber}}
}

@@

\begin{frame}{The goal of this talk}
    \begin{center}
        Write your interpreter by deforesting the object language,
        to~exhibit more static safety in a simpler type system.
    \end{center}
\end{frame}

@@

\onslide<1->
\begin{tabular}{@{}Mll@{}}
\only<5-8,11-14>{\text{\textcolor{structure.fg}{Object term}}}%
\only<16>{\rlap{\textcolor{structure.fg}{ML functor}}}
& \only<5-8,11-15>{\textcolor{structure.fg}{\only<-14>{\llap{\tikz\draw[->](0,0)--(1.5,0)node[anchor=mid,inner sep=0pt]{\vphantom{O}};\:}}\only<3-8>{Haskell term}\only<9-15>{ML functor}}} \\
\multicolumn{2}{@{}l@{}}{\only<16>{\alert{\texttt{module POWER (S:Symantics) = struct open S}}}} \\
\only<16>{\rlap{\alert{\texttt{ let term () =}}}}%
\visible<1-14>{\uncover<1-2,5-8,11-14>{\alert<6,12>{\fun{x} \fix{f} \fun{n}}}}
& \ttfamily
  \only<5-8>{\alert<6>{lam (\textbackslash x -> fix (\textbackslash f -> lam (\textbackslash n ->}}%
  \only<11-16>{\alert<12>{lam\,(fun x\,-> fix\,(fun f\,-> lam\,(fun n\,->}} \\
\visible<1-14>{\uncover<1-2,5-8,11-14>{\alert<8,14>{\cond{\textcolor{black}{n\le\alert<5,11>{0}}}{\textcolor{black}{1}}}}}
& \ttfamily
  \only<5-8>{\alert<8>{if\textunderscore} \alert<8>(leq n (\alert<5>{int 0})\alert<8>) \alert<8>(int 1\alert<8>)}%
  \only<11-16>{\alert<14>{if\textunderscore} \alert<14>(leq n (\alert<11>{int 0})\alert<14>) \alert<14>{(fun\,()\,->}\,int 1\alert<14>)} \\
\visible<1-14>{\uncover<1-2,5-8,11-14>{x\times \alert<7,13>{f(n-1)}}}
& \ttfamily
  \only<5-8>{\alert<8>(mul x (\alert<7>{app f (add n (int (-1)))})\alert<8>))))}%
  \only<11-16>{\alert<14>{(fun\,()\,->}\,mul x\,(\alert<13>{app f\,(add n\,(int\,(-1)\!)\!)}\!)\!\alert<14>)\!)\!)\!)} \\
\multicolumn{2}{@{}l@{}}{\only<16>{\alert{\texttt{end:\ functor (S:Symantics) -> sig}}}} \\
\only<16>{\rlap{\alert{\texttt{ val term:\ unit\,->}}}}%
\visible<1-14>{\uncover<1-2,5-8,11-14>{\mathord:\;\ZZ\to\ZZ\to\ZZ}}
& \ttfamily
  \only<5-8>{:\!:\ \!Symantics \!repr \!=>\,repr\,(Int\,->\,Int\,->\,Int)}%
  \only<11-16>{('c,\,int\,->\,int\,->\,int)\,\only<16>{\alert{S.}}repr} \\
\multicolumn{2}{@{}l@{}}{\only<16>{\alert{\texttt{end}}}}
\end{tabular}
\end{frame}

@@

\begin{frame}[fragile]{\only<11->{Type-indexed types}\only<13->{: }\only<-10,13-15>{Partial evaluation}\only<16->{CPS transformation}}
\setbeamercovered{transparent=20}
\newcommand\remember[1]{\tikz[remember picture,baseline=0pt,text=structure.fg] \node[inner ysep=0pt,inner xsep=2pt,anchor=base] (#1) {#1};}
\newcommand\remark[1]{\textcolor{structure.fg}{\footnotesize(* #1 *)}}
\definecolor{dynamic}{rgb}{.7,0,.7}
\definecolor{static}{rgb}{0,.7,.7}
\newcommand\dynamic[1]{\textcolor{dynamic}{\texttt{‹#1›}}}
\newcommand\static[1]{\textcolor{static}{\texttt{#1}}}
\begin{overlayarea}{0pt}{3ex}
\onslide+<1,13-15>{\rlap{\hspace*{-1pc}\texttt{module P = struct}}}%
\onslide+<12>\vspace*{-1em}%
\begin{semiverbatim}
module type Symantics = sig type ('c,\alert<12>{'s},'a) repr
 val int: int -> ('c,\alert<12>{int},int) repr
 val lam: 'x\,-> ('c, \alert<12>{('c,'s,'a) repr ->
            \,        ('c,'t,'b) repr} as\,'x, 'a\,->\,'b) repr
\uncover<0>{ val fix: (('c, \alert<12>{('c,'s,'a) repr -> ('c,'t,'b) repr},
                'a -> 'b) repr as 'x -> 'x) -> 'x}
 val app: ('c, \alert<12>{('c,'s,'a) repr ->
               ('c,'t,'b) repr} as\,'x, 'a\,->\,'b) repr ->\,'x
\uncover<0>{ val add: 'x -> 'x -> (('c,\alert<12>{int},int) repr as\,'x)
 val if_: ('c,\alert<12>{bool},bool) repr -> (unit->'x) -> (unit->'x)
                              -> (('c,\alert<12>{'s},'a) repr as\,'x)}
end
\end{semiverbatim}
\end{overlayarea}
\begin{tabular}[L]{@{}>{\ttfamily\frenchspacing}l@{\hspace*{-.5em}}r@{ }cc@{}}
\onslide+<1-10,13->
type ('c,\alt<13->{\only<14->{\alert<14-15>{'s},}'a}{\alt<1>{ 'a}{\alert<2>{int}}}) repr
                    & \onslide+<3-10>\remember{source term}
                    & \onslide+<4-10>$3$
                    & \onslide+<5-10>$x$ \\
\onslide+<1-10,13->
\ \ \ = \alt<13->{\alt<16->{\makebox[14pc][l]{\alert{('s\,->\,ans)\,->\,ans}}\remark{CBN CPS evaluator}}{('c,'a) code}}
                 {\alt<1>{???}{('c,int) code}}
                    & \onslide+<3-10>\remember{dynamic part}
                    & \onslide+<4-10>\dynamic{3}
                    & \onslide+<5-10>\dynamic{x} \\
\onslide+<2-10,13->
\ \ \ \alt<16->{= \makebox[14pc][l]{\alert{('c, ('s\,->\,ans)\,->\,ans) code}}\remark{CBN CPS compiler}}{* \alt<13->{\alert{\alt<14->{'s}{('c,'a) static}}}{int} option}
                    & \onslide+<3-10>\remember{static part}
                    & \onslide+<4-10>\static{Some 3}
                    & \onslide+<5-10>\static{None} \\
\onslide+<14-15>...\\
\onslide+<14-15>\rlap{\hspace*{-1pc}end}%
\onslide+<6-10>
type ('c,\alert<6>{int->int}) repr
                    && \onslide+<8-10>\hspace*{-4em}$\fun{x}x$
                    & \onslide+<7-10>$f$ \\
\onslide+<6-10>
\ \ \ = ('c,int->int) code
                    && \onslide+<8-10>\hspace*{-4em}\dynamic{fun \!x->x}
                    & \onslide+<7-10>\dynamic{f} \\
\onslide+<15>\rlap{\hspace*{-1pc}module POWER7P = POWER7(P)}%
\onslide+<6-10>
\ \ \ * \alert<8>{(('c,int) repr ->}
                    && \onslide+<8-10>\hspace*{-4em}\static{Some (fun \!r->r)}
                    & \onslide+<7-10>\static{None} \\
\onslide+<15>\rlap{\hspace*{-1pc}\>POWER7P.term ()}%
\onslide+<6-10>
\ \ \ \ \ \ \alert<8>{('c,int) repr)} option \\
\onslide+<15>\rlap{\hspace*{-1pc}\ \ (‹fun x -> x\,*\,x\,*\,x\,*\,x\,*\,x\,*\,x\,*\,x›, Some <fun>)}%
\\
\onslide+<9-10>
type ('c,'a) repr \\
\ \ \ = ('c,'a) code \\
\ \ \ * \alert<9-10>{\alt<9>{???}{('c,'a) static}} option
\end{tabular}
\onslide+<10-13,14->
\texttt{\begin{tabular}[L]{@{}l@{ }l@{}}
    type ('c,\:int)        &static = int \\
    type ('c,\:bool)       &static = bool \\
    type ('c,\:'a\,->\,'b) &static = ('c,'a)\:repr -> ('c,'b)\:repr
\end{tabular}}

@@

\begin{frame}{CPS transformations}
\begin{center}
\begin{tikzpicture}[auto]
    \node at (0,2) [lang] (lang1) {Simply typed $\lambda$-calculus};
    \node at (0,0) [lang] (lang2) {Simply typed $\lambda$-calculus};
    \draw [->] (lang1) to [bend right=30] node [swap] {CBV} (lang2);
    \draw [->] (lang1) to [bend left=30]  node        {CBN} (lang2);
    \draw [->] (lang2) -- +(-135:2cm) node [inner sep=0pt, anchor=north east] {evaluate};
    \draw [->] (lang2) -- +(-105:2cm) node [inner xsep=0pt, anchor=north east] {compile\hspace*{-.75em}};
    \draw [->] (lang2) -- +(-75:2cm)  node [inner xsep=0pt, anchor=north west] {\hspace*{-.75em}specialize};
    \draw      (lang2)    +(-45:2cm)  node {\dots};
\end{tikzpicture}
\end{center}

    Payoffs: evaluation order independence, mutable state
\end{frame}

\begin{frame}[fragile]{Other benefits}
\setbeamercovered{transparent=20}
\begin{block}{Supports initial type-checking}
    Type-check once, even under~$\lambda$, then interpret many times.

\begin{semiverbatim}
  FilePath -> \alert{Maybe} (exists a. Typeable a =>
                     forall repr. Symantics repr =>
                     repr a)
\end{semiverbatim}

    ``Typing dynamic typing'' (ICFP 2002) works.  We have the code.
\end{block}

\pause
\begin{block}{Preserves sharing in the metalanguage}
    Compute the interpretation of a repeated object term once, then use it many
    times.

    \smallskip
    $2\times3 + 2\times3$\hfill\texttt{let n = mul (int 2) (int 3) in add n n}
\end{block}

\pause
\begin{block}{Embed one object language in another}
\texttt{(Symantics repr, Symantics' repr') => repr (repr' Int)}\hspace*{-1em}
\end{block}
\end{frame}

\begin{frame}{Conclusion}
    Write your interpreter by deforesting the object language
    \begin{itemize}
    \item An abstract data type family
    \item Type-indexed types
    \end{itemize}
    Exhibit more static safety in a simpler type system
    \begin{itemize}
    \item Early, obvious guarantees
    \item Supports initial type-checking
    \item Preserves sharing in the metalanguage
    \item Embed one object language in another
    \end{itemize}
\end{frame}
